A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-11 tex/example.tex
[ tex ] Minor changes

\section{Examples}\label{sec:example}
Since our theory is not confined to a specific simply-typed language, we can employ it with any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings
\item spine calculus: a simply-typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first serves as a standard exercise in implementing a type synthesiser and the second for an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This also applies to parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:

\bgroup
\small
  \centering
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \and
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with the following operations to correspond with each rule:
\begin{align*}
  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{align*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can apply the type synthesiser presented in \Cref{sec:type-synthesis} directly to this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}

A spine calculus is a calculus with applications in spine form $t\;u_1\;\ldots\;u_n$ which consists of a head $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
This example might initially appear impossible, since the number of arguments a term construct specified by a signature can take  is finite.
However, we do not have any restriction on the number of symbols that may be inhabited in a signature, enabling us to have a corresponding type construct and term construct for each number $n$ of arguments.
This underscores the importance and necessity of having an arbitrary set (with decidable equality) as a set of operation symbols.
To be precise, consider the following rule

\bgroup
\small
  \centering
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn \textstyle{\prod_{i = 1}^{n}} A_i \bto B \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
extending the type signature $\Sigma_{\bto}$ with $\mathsf{fun}_n$ such that $\arity({\mathsf{fun}_n}) = n$ and the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with $\mathtt{app}_n$ such that 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}.
\]
Then each spine application $t\;u_1\;\ldots\;u_n$ of the length $n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

>>>>

\section{Examples}\label{sec:example}
Since our theory is not confined to a specific simply-typed language, we can employ it with any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings
\item spine calculus: a simply-typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This also applies to parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can apply the type synthesiser presented in \Cref{sec:type-synthesis} directly to this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}
A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of applications, as does its reflected syntax used for metaprogramming.

At first glance, accommodating this form of application may seem impossible, given the number of arguments a construct can accept is finite as specified by a signature.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn \textstyle{\prod_{i = 1}^{n}} A_i \bto B \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
The type signature $\Sigma_{\bto}$ can be extended with $\mathsf{fun}_n$ where $\arity({\mathsf{fun}_n}) = n$ for each $n : \N$.
Similarly, the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
also for each $n : \N$.
Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

### 2023-07-11 tex/sec1-intro.tex
[ tex ] Minor changes

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode}:
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{enumerate*}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does.
Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we present a first theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.

@@

Our work is partly inspired by \varcitet{Wadler2022}{'s} textbook, where programming language concepts are formally introduced using the proof assistant~\Agda based on Martin-L\"of Type Theory, and one particular topic is bidirectional type synthesis for~\PCF.
Their treatment deviates from existing literature:
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the simultaneously computational and logical nature of Martin-L\"of type theory to formulate algorithmic soundness, completeness, and decidability in one go.
%(although they do not emphasise the difference of their approach from the traditional one)
Recall that the law of excluded middle $P + \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether they are equal and gives an identity proof explicitly if they are, which may or may not be possible depending on the domain of $x$~and~$y$.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ has a derivation for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a contradiction proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Then, both algorithmic soundness and completeness in original form are implied.
%echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
%\todo{traditional completeness is still derivable from this `LEM'}
Our work takes the same approach but expands the scope significantly.
%That is, our decidable bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.

>>>>

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{enumerate*}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{enumerate*}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does.
Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.

@@

Our work is partly inspired by \varcitet{Wadler2022}{'s} textbook, where programming language concepts are formally introduced using the proof assistant~\Agda based on Martin-L\"of Type Theory, and one particular topic is bidirectional type synthesis for~\PCF.
Their treatment deviates from existing literature:
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the simultaneously computational and logical nature of Martin-L\"of type theory to formulate algorithmic soundness, completeness, and decidability in one go.
%(although they do not emphasise the difference of their approach from the traditional one)
Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in Martin-L\"of type theory and we say $P$ is \emph{decidable} if it holds for $P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether they are equal and gives an identity proof explicitly if they are, which may or may not be possible depending on the domain of $x$~and~$y$.
Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ has a derivation for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a contradiction proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Then, both algorithmic soundness and completeness in original form are implied.
%echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
%\todo{traditional completeness is still derivable from this `LEM'}
Our work takes the same approach but expands the scope significantly.
%That is, our decidable bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.

### 2023-07-11 tex/sec5-synthesis.tex
[ tex ] Minor changes

The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
Here we shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating parameter': a substitution $\rho_0$ \emph{partially defined} on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesis argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.

>>>>

The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
Here we shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesis argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.
